---
output:
  html_document:
        theme: readable
editor_options: 
  chunk_output_type: console
---

```{r}

library(tidyverse)

df <- read_rds(file = "ALS_progression_rate.1822x370.rds")  %>% tbl_df()

df <- df %>% rename(response = dFRS)

head(df)

```

# Split into training and prediction set

```{r}

set.seed(0)

data_train    <- df %>% filter(!is.na(response))
data_predict  <- df %>% filter(is.na(response))                              

dim(data_predict)
dim(data_train)

```

# My code

```{r}
library(tidyverse)
library(glmnet)
library(caret)

# Cargar datos
df <- read_rds(file = "ALS_progression_rate.1822x370.rds") %>% tbl_df()
df <- df %>% rename(response = dFRS)

# Dividir datos en entrenamiento y predicci??n
set.seed(0)
data_train    <- df %>% filter(!is.na(response))
data_predict  <- df %>% filter(is.na(response))

# Dividir en conjunto de entrenamiento (80%) y prueba (20%)
set.seed(0)
trainfold <- data_train %>% sample_frac(size = 0.80)
testfold  <- setdiff(data_train, trainfold)

# Extraer matriz de caracter??sticas y vector respuesta
x_train <- model.matrix(response ~ ., trainfold)[,-1]
y_train <- trainfold$response
x_test  <- model.matrix(response ~ ., testfold)[,-1]
y_test  <- testfold$response

# Definir funci??n para calcular RMSE
eval_rmse <- function(model, x, y) {
  preds <- predict(model, newx = x, s = "lambda.min")
  sqrt(mean((y - preds)^2))
}

# Ridge Regression (alpha = 0)
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0)
rmse_ridge <- eval_rmse(cv_ridge, x_test, y_test)

# Lasso Regression (alpha = 1)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1)
rmse_lasso <- eval_rmse(cv_lasso, x_test, y_test)

# Elastic Net (alpha = 0.5)
cv_elastic <- cv.glmnet(x_train, y_train, alpha = 0.5)
rmse_elastic <- eval_rmse(cv_elastic, x_test, y_test)

# Comparar RMSEs
rmse_results <- tibble(
  Model = c("Ridge", "Lasso", "Elastic Net"),
  RMSE  = c(rmse_ridge, rmse_lasso, rmse_elastic)
)

print(rmse_results)


```




# Plot of testfold observed and predicted values and residuals
 



```{r}
library(tidyverse)
library(glmnet)
library(caret)

# Cargar datos
df <- read_rds(file = "ALS_progression_rate.1822x370.rds") %>% tbl_df()
df <- df %>% rename(response = dFRS)

# Dividir datos en entrenamiento y predicci??n
set.seed(0)
data_train    <- df %>% filter(!is.na(response))
data_predict  <- df %>% filter(is.na(response))

# Dividir en conjunto de entrenamiento (80%) y prueba (20%)
set.seed(0)
trainfold <- data_train %>% sample_frac(size = 0.80)
testfold  <- setdiff(data_train, trainfold)

# Extraer matriz de caracter??sticas y vector respuesta
x_train <- model.matrix(response ~ ., trainfold)[,-1]
y_train <- trainfold$response
x_test  <- model.matrix(response ~ ., testfold)[,-1]
y_test  <- testfold$response

# Definir funci??n para calcular RMSE
eval_rmse <- function(model, x, y) {
  preds <- predict(model, newx = x, s = "lambda.min")
  sqrt(mean((y - preds)^2))
}

# Ridge Regression (alpha = 0)
cv_ridge <- cv.glmnet(x_train, y_train, alpha = 0)
rmse_ridge <- eval_rmse(cv_ridge, x_test, y_test)

# Lasso Regression (alpha = 1)
cv_lasso <- cv.glmnet(x_train, y_train, alpha = 1)
rmse_lasso <- eval_rmse(cv_lasso, x_test, y_test)

# Elastic Net Regression (alpha = 0.5)
cv_elastic <- cv.glmnet(x_train, y_train, alpha = 0.5)
rmse_elastic <- eval_rmse(cv_elastic, x_test, y_test)

# Mostrar los RMSE para los tres modelos
cat("RMSE for Ridge: ", rmse_ridge, "\n")
cat("RMSE for Lasso: ", rmse_lasso, "\n")
cat("RMSE for Elastic Net: ", rmse_elastic, "\n")

# Predicci??n y creaci??n de tibble con valores observados, predichos y residuales para los tres modelos
pred_ridge <- predict(cv_ridge, newx = x_test, s = "lambda.min")
pred_lasso <- predict(cv_lasso, newx = x_test, s = "lambda.min")
pred_elastic <- predict(cv_elastic, newx = x_test, s = "lambda.min")

# Crear tibble con los valores observados, predichos y residuales
pd <- tibble(
  observed = y_test,
  pred_ridge = pred_ridge,
  pred_lasso = pred_lasso,
  pred_elastic = pred_elastic
) %>%
  pivot_longer(cols = starts_with("pred_"), 
               names_to = "model", 
               values_to = "predicted") %>%
  mutate(residual = observed - predicted)

# Gr??fico combinado de observados vs predichos para los tres modelos con diferentes colores
p1 <- ggplot(pd, aes(x = predicted, y = observed, color = model)) +
  geom_point() +
  theme_classic() +
  labs(title = "Observed vs Predicted for Ridge, Lasso, and Elastic Net") +
  scale_color_manual(values = c("red", "blue", "green")) +
  theme(legend.title = element_blank())

# Gr??fico combinado de residuos vs valores predichos para los tres modelos con diferentes colores
p2 <- ggplot(pd, aes(x = predicted, y = residual, color = model)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_classic() +
  labs(title = "Residuals vs Predicted for Ridge, Lasso, and Elastic Net") +
  scale_color_manual(values = c("red", "blue", "green")) +
  theme(legend.title = element_blank())

# Mostrar ambos gr??ficos
print(p1)
print(p2)

```



# Predict the real unknown data

First we fit the model to all of our known data

Then we predict on the unknown data

The predictions must have the following column and the row order must be the same as the original!

* predicted (the predicted value)


```{r}





xx_train <- model.matrix(response ~ ., data_train)[,-1]
yy_train <- data_train$response

xx_test <- as.matrix(data_predict[, -1])

dim(xx_train)
dim(xx_test)
cv_lasso <- cv.glmnet(xx_train, yy_train, alpha = 1)

predicted <- predict(cv_lasso, newx = xx_test, s = "lambda.min")



submission <- tibble(predicted)

head(submission)

```


# Submitting your answer

The following code will give us

* your chosen team name
* the name of the people on the team
* your estimated RMSE (from train/test or CV or similar)
* your predictions

Please edit the values below .

The filename of the output will be automated als_progression.TEAMNAME.rds

Please - do not use space or funny letters in your team name.

```{r}

team_name        <- "team_pauet"
team_people      <- c("Pablo_Bondia")
team_error_rate  <- test_rmse
team_predictions <- submission # This should be a tibble with a column called "predicted"

#
# Always run this code
# If it fails you have done something wrong!
#
# Extract the columns needed
team_predictions <- team_predictions %>% select(predicted)

# Save all the stuff in one object
write_rds(x = list(team_name, team_people, team_error_rate, team_predictions), 
          file = paste("als_progression.", team_name, ".rds", sep=""))

```

# Checking format of all saved objects

```{r}

files   <- Sys.glob("als_progression.*.rds")
results <- tibble(filename = files, team_name=NA, team_people=NA, team_rmse=NA,n=NA, mean=NA)

for (i in 1:nrow(results)) {
  x <- read_rds(file = as.character(results$filename[i]))
  results$team_name[i]        <- x[[1]]
  results$team_people[i]      <- paste(x[[2]], collapse=",", sep=" ")
  results$team_rmse[i]        <- x[[3]]
  y                           <- x[[4]]
  results$n                   <- nrow(y)
  results$mean                <- mean(y$predicted, na.rm = T)
  results$submission[i]       <- list(x[[4]])
}

rm(x,y)

results %>% select(-filename)

results$submission[[1]]

```

# Upload your rds file!


