---
title: "Round3"
author: "Pablo Bondia"
date: "2025-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Load necessary libraries
library(tidyverse)
library(caret)  # For model training
library(glmnet)  # For elastic net
library(ranger)  # For Random Forest
library(gbm)  # For Gradient Boosting
library(e1071)  # For SVM
library(Metrics)  # For RMSE evaluation

# Load the data
df <- read_rds(file = "ALS_progression_rate.1822x370.rds")  %>% tbl_df()

# Rename the response variable
df <- df %>% rename(response = dFRS)

# Display the first few rows of the data
head(df)

```

```{r}
# Split data into training and prediction sets
data_train    <- df %>% filter(!is.na(response))
data_predict  <- df %>% filter(is.na(response))

# Check dimensions of both sets
dim(data_predict)
dim(data_train)

```

```{r}
# Create training and test folds (80-20 split)
set.seed(0)
train_indices <- createDataPartition(data_train$response, p = 0.8, list = FALSE)
trainfold <- data_train[train_indices, ]
testfold <- data_train[-train_indices, ]

# Create model matrices (without the response column)
x_train <- model.matrix(response ~ ., trainfold)[,-1]
x_test <- model.matrix(response ~ ., testfold)[,-1]

# Convert response variable to numeric vectors
y_train <- trainfold$response
y_test <- testfold$response

```

DATAFRAME EXPLORATION: PCA

If you see that only a small number of principal components explain most of the variance (for example, the first 2-3 components explain 80% or more of the variance), that’s a good indicator that there are underlying correlations between the variables. This is because correlated variables typically collapse into a few principal components.

```{r}
# Select numeric columns for PCA
numeric_data <- data_train %>% select_if(is.numeric)

# Scale and center the data before applying PCA
pca <- prcomp(numeric_data, scale. = TRUE)

# Summary of PCA to see variance explained
summary(pca)

# Extract variance explained
var_explained <- summary(pca)$importance[2, ]  # Second row is the proportion of variance explained

# Plot the variance explained by each component
library(ggplot2)

# Create a data frame for plotting
pca_variance_df <- data.frame(
  Principal_Component = 1:length(var_explained),
  Variance_Explained = var_explained
)

# Scree plot
ggplot(pca_variance_df, aes(x = Principal_Component, y = Variance_Explained)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Variance Explained by Each Principal Component",
       x = "Principal Component", 
       y = "Variance Explained") +
  theme_minimal()

# Plot the cumulative variance explained
cumulative_variance <- cumsum(var_explained)

ggplot(data.frame(Principal_Component = 1:length(cumulative_variance), 
                  Cumulative_Variance = cumulative_variance), 
       aes(x = Principal_Component, y = Cumulative_Variance)) +
  geom_line(color = "darkred", size = 1.2) +
  geom_point(color = "darkred") +
  labs(title = "Cumulative Variance Explained by Principal Components",
       x = "Principal Component", 
       y = "Cumulative Variance Explained") +
  theme_minimal()

```






```{r}
# Step 3: Find the number of components that explain 97.5% of the variance
threshold <- 0.99
n_components <- which(cumulative_variance >= threshold)[1]

# Print the number of components and the explained variance
cat("Number of components that explain at least 99% of the variance:", n_components, "\n")
cat("Cumulative variance explained by the selected components:", cumulative_variance[n_components], "\n")
```


```{r}
#Step 4: Apply PCA to transform both training and test datasets based on the selected components
# Make sure to select only the numeric features from the train and test sets
numeric_train <- x_train[, colnames(x_train) %in% colnames(numeric_data)]
numeric_test <- x_test[, colnames(x_test) %in% colnames(numeric_data)]

# Perform PCA transformation on both training and test data
x_train_pca <- predict(pca, newdata = numeric_train)[, 1:n_components]
x_test_pca <- predict(pca, newdata = numeric_test)[, 1:n_components]

# Check the dimensions of the reduced data
cat("Training data dimensions after PCA:", dim(x_train_pca), "\n")
cat("Test data dimensions after PCA:", dim(x_test_pca), "\n")

```










```{r}
# Set up training control for repeated cross-validation with 3 repetitions
trControl <- trainControl(method = "repeatedcv", number = 10, 
                          repeats = 3,  # Repeat 3 times (can adjust)
                          verboseIter = TRUE)

# Initialize result tibble
result <- tibble(method = c("glmnet", "ranger", "gbm", "svmRadial"),
                 accuracy_train = NA, 
                 accuracy_test = NA,
                 runtime = NA)


```


```{r}
# Loop through each model
for(i in 1:nrow(result)) {
  
  # Define the method to use
  method <- result$method[i]
  
  # Log the model being trained
  cat(paste(Sys.time(), "Doing", method, "\n"), file = "log.txt", append = TRUE)
  
  # Record start time for runtime calculation
  starttime <- as.integer(Sys.time())
  
  # Fit the model using caret
  fit <- train(x = x_train, 
               y = y_train, 
               method = method,
               tuneLength = 10,  # Tune hyperparameters
               trControl = trControl)
  
  # Calculate runtime
  runtime <- as.integer(Sys.time()) - starttime
  
  # Calculate RMSE for test data (regression task)
  test_predictions <- predict(fit, newdata = x_test)
  test_rmse <- rmse(y_test, test_predictions)  # RMSE evaluation

  # Save performance and runtime
  result$accuracy_train[i] <- mean(fit$resample$RMSE)  # For regression, using RMSE (Root Mean Squared Error)
  result$accuracy_test[i] <- test_rmse
  result$runtime[i] <- runtime
  
  # Log model completion
  cat(paste(Sys.time(), "Done", method, "Runtime:", runtime, "\n"), file = "log.txt", append = TRUE)
}

# Print the results of model comparisons
print(result)

```

Plan: 

1- Check for colinearity. TO SEE WHAT TO USE AND MAYBE DO REGULARIZATION OR PCA. Remember elastic net (lasso part) is sensitive to colinearity, also sv. USE CORRELATION MATRIX OR VIF OR PCA (see if a lot of the variance is explained by few PCs)

Didnt find a lot of colinearity.

2-Check for outliers, remove them and make models that become very good at normal data and bad at outliers.


3- Check to see if diff models capture diff aspects of the date? TO SEE IF I SHOULD AVERAGE PREDICITONS OF DIFFERENT MODELS. Metalearning is another option but i dont like it

3- Look at skewed data?

4- Feature engineering?




4- TEST IF THE MODELS AR SIGN DIFFERENT, USING ANOVA.

Model Type	Recommended Dimensionality Reduction Techniques
Linear Models (ElasticNet, Lasso, etc.)	PCA (to handle multicollinearity and reduce noise)
Tree-Based Models (Random Forest, GBM)	None (Tree models can handle high-dimensional and correlated data well)
Non-Linear Models (SVM, Neural Networks)	PCA (if linear relationships exist), UMAP (for non-linear relationships)
Clustering Models (K-means, DBSCAN)	PCA (for linear clusters), UMAP (for non-linear clusters)





```{r}

```

